Correctly Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
Gentler intro to frontier-based semantics.

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \maddr \mbox{ where } \maddr \notin \msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie,
$\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$ and $\mstate.\msto$ is notation for projecting out the
store component of a state. These functions appear to ignore $\mlab$
and $\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \widehat{\State} \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\\
\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
We associate the value in hand with the expression that produced
it. Also, we fan out to all functions that are associated with the
original function position, because applied functions are in
\emph{strict position} (more about that later).

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic ``worklist''
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
              S' &= \setof{(c,\msto') \mid c \in F'} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{align*}
\Sigma &\in \Store^* \\
S &\subseteq {\mathbb N} \times \widehat{\State} \\
F &\subseteq \widehat{\State} \\
(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
              S' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{align*}
\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
 S' &= \setof{(\mastate,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}

@@

\begin{lemma}
$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.

@@

\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\msto,\mval) &= \setof{\mval}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\maddr_\mkont),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.

>>>>

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
\end{align*}
These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
used to determinize the choice of fresh addresses. The $\sqcup$ on
stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
relation is non-deterministic in its choice of addresses, however it
must always choose a fresh address when allocating a continuation or
variable binding.  If we consider machine states equivalent up to
consistent renaming and fix an allocation scheme, this relation
defines a deterministic machine (the relation is really a function).

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

HERE

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\mkaddr}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision.

\begin{align*}
\co[^\mcntr]{\kfn{\maddr_f,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is demanded in order to continue with
evaluation. More generally, this is because applied functions are in
\emph{strict position} (more about that later).

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic worklist
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

The state space changes from one store plus a set of contexts to a set
of seen abstract states (context plus store), $S$, a set of contexts to
step (the frontier), $F$, and the store to step those contexts with, $\msto$.

\begin{equation*}
(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
\end{equation*}

We constantly see more states, so $S$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):

\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}

The next store is the least upper bound of all the stores in $I$ (in
order to have only one, and keep the semantics sound):

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}

The next frontier is exactly the states that we found from stepping the last frontier, but have not seen before. They must be \emph{states}, so we pair the contexts with the next store:

\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
\end{equation*}

Finally, we add what we know we had not yet seen to the seen set:

\begin{equation*}
  S' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}

To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{align*}
\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
\end{align*}
\begin{align*}
(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              (t',\Sigma') &=\left\{\begin{array}{ll}
                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
                           (t,\Sigma)   & \text{ otherwise}
                          \end{array}\right. \\
              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
              S' &= \setof{(c,t') \mid c \in F'}\\
\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{align*}
\dmachstep &\subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) \\
\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
 S' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}

@@

\begin{lemma}
$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.

@@

\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\msto,\mval) &= \setof{\mval}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\mkaddr),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.
